\textbf{Evaluation of machine learning methods for 16S rRNA gene data}

### 2019-05-15 submission/manuscript.tex
modulate and change

Advances in sequencing technology and decreasing costs of generating 16S
rRNA gene sequences have allowed rapid exploration of human associated
microbiome and its health implications. Currently, the human microbiome
field is growing at an unprecedented rate and as a result, there is an
increasing demand for methods that identify associations between members
of the microbiome and human health. However, this is an undertaking as
human associated microbial communities are remarkably complex and
uneven. It is unlikely that a single species can explain a disease.
Instead, subsets of those communities, in relation to one another and to
their host, account for the differences in health outcomes. Machine
learning (ML) methods are effective at recognizing and highlighting
patterns in complex microbial datasets. Therefore, researchers have
started to explore the utility of ML models that use microbiota
associated biomarkers to predict human health and to understand the
microbial ecology of diseases such as liver cirrhosis, colorectal
cancer, inflammatory bowel diseases (IBD), obesity, and type 2 diabetes
(1--11). However, currently the field's use of ML lacks clarity and
consistency on which methods are used and how these methods are
implemented. Moreover, there is a lack of deliberation on why a
particular ML model is utilized. Recently, there is a trend towards
using more complex ML models such as random forest, extreme gradient
boosting and neural networks without a discussion on if and how much
model interpretibility is necessary for the study (11--14). The lack of
transparency on modeling methodology and model selection negatively
impact model reproducibility and reliability. We need to strive toward
better machine learning practices by (1) implementing consistent,
accessible and transparent machine learning methods; (2) selecting ML
models that reflect the goal of the study as it will inform our
expectations of model accuracy, complexity, interpretibility and
computational efficiency.

To showcase transparent ML methodologies and to shed light on how much
ML model selection can affect modeling results, we performed an
empirical analysis comparing several different ML models using the same
dataset. We used a previously published colorectal cancer (CRC) study
(3) which had fecal 16S rRNA gene sequences and human hemoglobin
concentrations from 490 patients. We built ML models using fecal 16S
rRNA gene sequences and human hemoglobin concentrations to predict
patients with normal colons or patients with colonic tumors which are
called screen relevant neoplasias (SRN). The study had 261 normal and
229 SRN samples. We established modeling pipelines for L2-regularized
logistic regression, L1 and L2 support vector machines (SVM) with linear
and radial basis function kernels, a decision tree, random forest and
XGBoost which are ML models that increase in complexity respectively.
Our ML pipeline utilized held-out test data to evaluate generalization
and prediction performance of each ML model. The mean test AUROC varied
from 0.73 (std ± 0.04) to 0.82 (std ± 0.03). Random Forest and XGBoost
had the highest mean AUROC for detecting SRN. Despite the simplicity,
the L1-regularized linear kernel SVM followed Random Forest and XGBoost
in performance. In terms of computational efficiency, L2 logistic
regression trained the fastest (0.001 hours, std ± 0.001), while XGBoost
took the longest (0.231 hours, std ± 0). We found that mean
cross-validation and testing AUROC varied only 0.007, which highlights
the importance of a seperate held-out test set and consistent
preprocessing/splitting data for evaluation. Aside from evaluating
generalization and classification performances for each of these models,
this study established standards for modeling pipelines of
microbiome-associated machine learning models.

@@

The data has 6921 features (6920 OTUs and FIT) and a two-class label
that defines the colonic health of the patients (SRN or normal). We
established modeling pipelines for binary classification with
L2-regularized logistic regression, L1 and L2 support vector machines
(SVM) with linear and radial basis function kernels, a decision tree,
random forest and XGBoost to emphasize the differences in model
accuracy, complexity, interpretibility and computational efficiency due
to model selection. We randomly split the data into stratified
training/validation and test sets 100 times with a 80/20 proportion
{[}Figure 1{]}. For each data-split, training/validation set consisted
of 393 patients (209 SRN), while the test set was composed of 97
patients (52 SRN). The training/validation data was used for training
purposes and validation of hyperparameter selection, and the test set
was used for evaluation purposes. Validation of hyperparameter selection
was performed by randomly splitting the training/validation set into
stratified training and validation sets 100 times with a 80/20
proportion {[}Figure 1{]}. For L2-regularized logistic regression, L1
and L2 support vector machines (SVM) with linear and radial basis
function kernels we tuned the \textbf{cost} hyperparameter which
determines the regularization strength where smaller values specify
stronger regularization. For SVM with radial basis function kernel we
also tuned \textbf{sigma} hyperparameter which determines the reach of a
single training instance where for a high value of sigma, the SVM
decision boundary will be dependent on the points that are closest to
the decision boundary. For the decision tree model, we tuned the
\textbf{depth of the tree} where deeper the tree, the more splits it
has. For random forest, we tuned the \textbf{number of features} to
consider when looking for the best tree split. For xgboost, we tuned for
\textbf{learning rate} and the \textbf{fraction of samples} to be used
for fitting the individual base learners. We validated the prediction
performances of each hyperparameter setting (Figure S1{]}.

@@

We evaluated the prediction performance of seven binary classification
models when applied to held-out test data over 100 data-splits using
AUROC as classification performance metric. Random Forest and XGBoost
had the highest mean AUROC for detecting SRN, 0.82 (std ± 0.033) and
0.814 (std ± 0.04) respectively {[}Figure 2{]}. L1 linear SVM had
significantly lower AUROC values, 0.76 (std ± 0.054) and 0.742 (std ±
0.04) {[}Figure 2{]}. However, it also had significanly higher
performances than L2 linear SVM, RBF SVM, L2 logistic regression and
decision tree which had mean AUROC values of 0.74 (std ± 0.045), 0.73
(std ± 0.041), 0.745 (std ± 0.039) and 0.742 (std ± 0.04), respectively
{[}Figure 2{]}. We also evaluated the generalization performance of each
classifier by comparing their mean cross-validation AUROC and mean
testing AUROC. We found that mean cross-validation and testing AUROC
difference for each model was below 0.01.

@@

We interpreted the feature weights of L1 and L2 SVM with linear kernel
and regression coefficients of L2 logistic regression using the training
data. To get a sense of the most important features in linear models, we
calculated the mean weights of all the features over the 100 data-splits
for each model. {[}Figure 4{]}. In the three linear models, FIT was the
feature that drove the detection of SRNs, followed by eight OTUs which
belong to family \emph{Lachnospiraceae} and \emph{Ruminococcaceae} (four
with positive signs; Otu01239, Otu00742, Otu00008 Otu00659, and four
with negative signs; Otu00015, Otu000150, Otu00609, Otu00629). To get
the importance of features in non-linear models, we

@@

Linear models trained faster than non-linear models. L2 logistic
Rregression and L1 and L2 SVM with linear kernel had training times of 0
min, (std ± 0), 13.8 min, (std ± 1.2) and 0 min, (std ± 0) respectively.
Whereas, SVM with radial basis function kernel, decision tree, random
forest and xgboost had training times of 0 hrs, (std ± 0), 0 hrs, (std ±
0), 0 hrs, (std ± 0) and 0 hrs, (std ± 0), respectively {[}Figure 4{]}.

@@

>>>>

Advances in sequencing technology and decreasing costs of generating 16S
rRNA gene sequences have allowed rapid exploration of human associated
microbiome and its health implications. Currently, the human microbiome
field is growing at an unprecedented rate and as a result, there is an
increasing demand for methods that identify associations between members
of the microbiome and human health. However, this is difficult as human
associated microbial communities are remarkably complex and uneven. It
is unlikely that a single species can explain a disease. Instead,
subsets of those communities, in relation to one another and to their
host, account for the differences in health outcomes. Machine learning
(ML) methods are effective at recognizing and highlighting patterns in
complex microbial datasets. Therefore, researchers have started to
explore the utility of ML models that use microbiota associated
biomarkers to predict human health and to understand the microbial
ecology of diseases such as liver cirrhosis, colorectal cancer,
inflammatory bowel diseases (IBD), obesity, and type 2 diabetes (1--11).
However, currently the field's use of ML lacks clarity and consistency
on which methods are used and how these methods are implemented. Most
notably, there are misleading practices of using ``leaky'' ML pipelines
where models are tested on training data or reporting the best outcome
of different iterations of cross-validation. Moreover, there is a lack
of discussion on why a particular ML model is utilized. Recently, there
is a trend towards using more complex ML models such as random forest,
extreme gradient boosting and neural networks without a discussion on if
and how much model interpretibility is necessary for the study (11--14).
The lack of transparency on modeling methodology and model selection
negatively impact model reproducibility and reliability. We need to
strive toward better machine learning practices by (1) implementing
consistent and correct machine learning pipelines; (2) selecting ML
models that reflect the goal of the study as it will inform our
expectations of model accuracy, complexity, interpretibility and
computational efficiency.

To showcase reliable ML methodologies and to shed light on how much ML
model selection can affect modeling results, we performed an empirical
analysis comparing several different ML models using the same dataset
and a robust ML pipeline. We used a previously published colorectal
cancer (CRC) study (3) which had fecal 16S rRNA gene sequences from 490
patients. We built ML models using fecal 16S rRNA gene sequences to
predict patients with normal colons or patients with colonic tumors
which are called screen relevant neoplasias (SRN). The study had 261
normal and 229 SRN samples. We established modeling pipelines for
L2-regularized logistic regression, L1 and L2 support vector machines
(SVM) with linear and radial basis function kernels, a decision tree,
random forest and XGBoost which are ML models that increase in
complexity respectively. Our ML pipeline utilized held-out test data to
evaluate generalization and prediction performance of each ML model. The
mean test AUROC varied from 0.598 (std ± 0.044) to 0.697 (std ± 0.037).
Random forest had the highest mean AUROC for detecting SRN. Despite the
simplicity, the L2-regularized logistic regression followed random
forest in performance. In terms of computational efficiency, L2 logistic
regression trained the fastest (0.202 hours, std ± 0.019), while XGBoost
took the longest (162.843 hours, std ± 10.018). We found that mean
cross-validation and testing AUROC varied only 0.01, which highlights
the importance of a seperate held-out test set and consistent
preprocessing of the data prior to evaluation. Aside from evaluating
generalization and classification performances for each of these models,
this study established standards for modeling pipelines of
microbiome-associated machine learning models.

@@

The data has 6920 features (6920 OTUs) and a two-class label that
defines the colonic health of the patients (SRN or normal). We
established modeling pipelines for binary classification with
L2-regularized logistic regression, L1 and L2 support vector machines
(SVM) with linear and radial basis function kernels, a decision tree,
random forest and XGBoost to emphasize the differences in model
accuracy, complexity, interpretibility and computational efficiency due
to model selection. We randomly split the data into stratified
training/validation and test sets 100 times with a 80/20 proportion
{[}Figure 1{]}. For each data-split, training/validation set consisted
of 393 patients (209 SRN), while the test set was composed of 97
patients (52 SRN). The training/validation data was used for training
purposes and validation of hyperparameter selection, and the test set
was used for evaluation purposes. Validation of hyperparameter selection
was performed by randomly splitting the training/validation set into
stratified training and validation sets 100 times with a 80/20
proportion {[}Figure 1{]}. For L2-regularized logistic regression, L1
and L2 support vector machines (SVM) with linear and radial basis
function kernels we tuned the \textbf{cost} hyperparameter which
determines the regularization strength where smaller values specify
stronger regularization. For SVM with radial basis function kernel we
also tuned \textbf{sigma} hyperparameter which determines the reach of a
single training instance where for a high value of sigma, the SVM
decision boundary will be dependent on the points that are closest to
the decision boundary. For the decision tree model, we tuned the
\textbf{depth of the tree} where deeper the tree, the more splits it
has. For random forest, we tuned the \textbf{number of features} to
consider when looking for the best tree split. For xgboost, we tuned for
\textbf{learning rate} and the \textbf{fraction of samples} to be used
for fitting the individual base learners. We validated the prediction
performances of each hyperparameter setting (Figure S1{]}.

@@

We evaluated the prediction performance of seven binary classification
models when applied to held-out test data over 100 data-splits using
AUROC as classification performance metric. Random Forest had the
highest mean AUROC for detecting SRN, 0.697 (std ± 0.037) {[}Figure
2{]}. L2 logistic regression and XGBoost had significantly lower AUROC
values, 0.676 (std ± 0.042) and 0.678 (std ± 0.04) {[}Figure 2{]}.
Random forest also had significanly higher performances than L2 linear
SVM, RBF SVM and decision tree which had mean AUROC values of 0.671 (std
± 0.045), 0.663 (std ± 0.044), and 0.598 (std ± 0.044), respectively
{[}Figure 2{]}. We also evaluated the generalization performance of each
classifier by comparing their mean cross-validation AUROC and mean
testing AUROC. We found that mean cross-validation and testing AUROC
difference for each model was below 0.01.

@@

We interpreted the feature weights of L1 and L2 SVM with linear kernel
and regression coefficients of L2 logistic regression using the training
data. To get a sense of the most important features in linear models, we
calculated the mean weights of all the features over the 100 data-splits
for each model. {[}Figure 4{]}. In the three linear models, FIT was the
feature that drove the detection of SRNs, followed by eight OTUs which
belong to family \emph{Lachnospiraceae} and \emph{Ruminococcaceae} (four
with positive signs; Otu01239, Otu00742, Otu00008, Otu00659, and four
with negative signs; Otu00015, Otu000150, Otu00609, Otu00629). To get
the importance of features in a decision tree, we used gini importance
which allows us to use the improvement in the split-criterion is the
importance measure attributed to the splitting variable

@@

Linear models trained faster than non-linear models. L2 logistic
Rregression and L1 and L2 SVM with linear kernel had training times of
13.2 min, (std ± 1.8), 563.4 min, (std ± 46.8) and 12.6 min, (std ± 1.8)
respectively. Whereas, SVM with radial basis function kernel, decision
tree, random forest and xgboost had training times of 101.32 hrs, (std ±
10.02), 162.84 hrs, (std ± 3.99), 0.2 hrs, (std ± 0.02) and 64.71 hrs,
(std ± 9.85), respectively {[}Figure 4{]}.

@@

The classification models were built to classify normal versus adenoma,
normal versus carcinoma, and high versus low SCFA concentrations. The
regression models were built to classify the SCFA concentrations of
acetate, butyrate, and propionate regardless of disease status. The
featThe bacterial abundances are the features used to predict colonic
disease status. Bacterial abundances are discrete data in the form of
Operational Taxonomic Unit (OTU) counts. There are 6920 OTUs for each
sample. FIT levels are continuous data present for each sample. Because
the data are in different scales, features are transformd by scaling
each feature to a {[}0-1{]} range

To evaluate the model performances, we randomly split the data into
training/validation and test sets 100 times with a 80/20 proportion.
Data-splits were stratified for diagnosis or SCFA concentration levels
in classification models. For each data-split, training/validation set
consisted of x patients (x carcinoma, y adenoma, z normal), while the
test set was composed of x patients (x carcinoma, y adenoma, z normal).
The training/validation data was used for training purposes and
validation of hyperparameter selection, and the test set was used for
evaluation purposes. Validation of hyperparameter selection was
performed by randomly splitting the training/validation set into
stratified training and validation sets 100 times with a 80/20
proportion {[}Figure 1{]}.
